---
title: "Trend and Seasonality"
author: "Luana Lima"
date: "03/04/2021"
output: pdf_document
---

## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

```

## Importing data

Let's continue working with our inflow data for reservoirs in Brazil. 

```{r}

#Importing time series data from text file#
raw_inflow_data <- read.table(file="../Data/inflowtimeseries.txt",header=FALSE,skip=0)

#Trim the table to include only columns you need
nhydro <- ncol(raw_inflow_data)-2
nobs <- nrow(raw_inflow_data) 

#If your file does not have header like this one you can add column names after 
#creating the data frame
colnames(raw_inflow_data)=c("Month","Year", "HP1", "HP2","HP3","HP4", "HP5",
                            "HP6","HP7", "HP8","HP9","HP10", "HP11","HP12", 
                            "HP13", "HP14","HP15")

#Checking data
head(raw_inflow_data)
str(raw_inflow_data)
```
## Creating the date object

Here we use the function my() from package lubridate.

```{r}
#using package lubridate
my_date <- paste(raw_inflow_data[,1],raw_inflow_data[,2],sep="-")
my_date <- my(my_date)  #function my from package lubridate
head(my_date)
#add that to inflow_data and store in a new data frame 
inflow_data <- cbind(my_date,raw_inflow_data[,3:(3+nhydro-1)])
head(inflow_data)

```


## Initial Plots

Initial time series plot.

```{r}
#using package ggplot2
for(i in 1:nhydro){
  print(ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+i)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+i)],sep="")) 
        )
}

```
## Zeros in the end on data 

The initial plots showed that we have zeros in the end of the data set. It could be missing observation or observation that haven't been observed yet. Use the tail() to find out how many zeros you have and how many lines you will need to remove.

```{r}
#check the final obs on data 
tail(inflow_data)


```
Note our last observation is from August 2011 but the data file was filled with zeros. Let's remove the last four rows of our data set.

```{r}
#Remove last for rows by replacing current data frame
inflow_data <- inflow_data[1:(nobs-4),]

#update object with number of observations
nobs <- nobs-4

#Tail again to check if the rows were correctly removed
tail(inflow_data)
```
Fixed!

## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_inflow_data <- ts(inflow_data[,2:(2+nhydro-1)],frequency=12) 
#note that we are only transforming columns with inflow data, not the date columns  #start=my_date[1],end=my_date[nobs],
head(ts_inflow_data,15)
```

Note that ts_inflow_data has information on start, end and frequency. 

## Plotting ACF and PACF

Let's use functions Acf() and Pacf() from package "forecast".

```{r acf}
#Acf and Pacf for HP1
for(i in 1:nhydro){
  par(mfrow=c(1,2))  #place plot side by side
  Acf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep="")) 
  # because I am not storing Acf() into any object, I don't need to specify plot=TRUE 
  Pacf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep=""))
}

```


## Trend Component

Let's identify and remove trend component like we leaned on the recorded videos for M4. You start by fitting a linear model to $Y_t=\beta_0+\beta_1*t+\epsilon_t$. 

```{r}
#Create vector t
t <- c(1:nobs)

#Choose one hydro plant to study, as an exercise try to generalize this routine for all 15 HP
#from the plot HP4 seems to have a trend so let's play with that column
iHP=4  #change this to chekc other HP
#prep_data <- data.frame("Inflow"=inflow_data[,iHP],"Time"=t)

#Fit a linear trend to TS of iHP
linear_trend_model=lm(inflow_data[,iHP+1]~t)  
summary(linear_trend_model)
beta0=as.numeric(linear_trend_model$coefficients[1])  #first coefficient is the intercept term or beta0
beta1=as.numeric(linear_trend_model$coefficients[2])  #second coefficient is the slope or beta1

#Let's plot the time series with its trend line
ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+iHP)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+iHP)],sep="")) +
            #geom_abline(intercept = beta0, slope = beta1, color="red")
            geom_smooth(color="red",method="lm") 

```


```{r}
#remove the trend from series
detrend_inflow_data <- inflow_data[,(iHP+1)]-(beta0+beta1*t)

#Understanding what we did
ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+iHP)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+iHP)],sep="")) +
            #geom_abline(intercept = beta0, slope = beta1, color="red")
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_inflow_data), col="green")+
            geom_smooth(aes(y=detrend_inflow_data),color="orange",method="lm") 

```
Note that blue line is our original series, red lien is our trend, green line is our original series minus the trend or in other words the detrended series. And in orange is the trend line for the detrended series which has slope 0 meaning we were able to effectively eliminate the trend with a linear model.

## Seasonal Component

Now let's shift attention to the seasonal component.

```{r}
#Let's choose another HP
iHP=1

#Use seasonal means model
#First create the seasonal dummies
dummies <- seasonaldummy(ts_inflow_data[,iHP])  
#this function only accepts ts object, no need to add one here because date 
#object is not a column

#Then fit a linear model to the seasonal dummies
seas_means_model=lm(inflow_data[,(iHP+1)]~dummies)
summary(seas_means_model)

#Look at the regression coefficient. These will be the values of Beta

#Store regression coefficients
beta_int=seas_means_model$coefficients[1]
beta_coeff=seas_means_model$coefficients[2:12]

#compute seasonal component
inflow_seas_comp=array(0,nobs)
for(i in 1:nobs){
  inflow_seas_comp[i]=(beta_int+beta_coeff%*%dummies[i,])
}

#Understanding what we did
ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+iHP)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+iHP)],sep="")) +
            geom_line(aes(y=inflow_seas_comp), col="red")


#Removing seasonal component
deseason_inflow_data <- inflow_data[,(1+iHP)]-inflow_seas_comp

#Understanding what we did
ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+iHP)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+iHP)],sep="")) +
            geom_line(aes(y=deseason_inflow_data), col="green")

```

