---
title: "Lab 4: Importing Data Sets"
author: "Luana Lima"
date: "02/10/2021"
output: pdf_document
---

## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

```

## Importing data with read.xlsx()

Importing data set without change the original file using read.xlsx

```{r}
library(xlsx)

energy_data1 <- read.xlsx(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow = 13,sheetIndex=1) #startRow is equivalent to skip on read.table

#Now let's extract the column names from row 11
read_col_names <- read.xlsx(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow = 11,endRow=11,sheetIndex=1)

colnames(energy_data1) <- read_col_names
head(energy_data1)
```

## Importing data with read_excel()

```{r}
library(readxl)

#Importing data set without change the original file using read.xlsx
energy_data2 <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Now let's extract the column names from row 11
read_col_names <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(energy_data2) <- read_col_names
head(energy_data2)

ts_energy <- ts(energy_data2[2:14])

```

## Importing data with read.csv()

```{r}
#no need to library additional packages

#Importing data set after manipulating original data file xlsx
energy_data3 <- read.csv(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", header = TRUE, stringsAsFactors = TRUE) 

#check that colnames are already stored because we deleted the units line so we were able to automatically import it
colnames(energy_data3)
head(energy_data3)
```

## Importing data with read.table()

```{r}
#no need to library additional packages

#Importing data set after manipulating original data file xlsx
energy_data4 <- read.table(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", header=TRUE,sep = ",",dec = ".") 

#check that colnames are already stored because we deleted the units line so we were able to automatically import it
colnames(energy_data4)
head(energy_data4)
```