---
title: "M10: State Space Models"
author: "Luana Lima"
date: "03/26/2021"
output: pdf_document
always_allow_html: true
editor_options: 
  chunk_output_type: inline
---

## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output.

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
#library(Kendall)
library(tseries)
#library(outliers)
library(tidyverse)
library(smooth)

#New package for M9 to assist with tables
#install.packages("kableExtra")
library(kableExtra)

```

## Importing data

For this module we will continue to work with the electricity retail price in US dataset from the U.S. Energy Information Administration. You may download [here][[https://www.eia.gov/electricity/data/browser/\#/topic/7?agg=2,0,1&geo=g&freq=M%2013:41:41%20GMT-0500%20(EST)](https://www.eia.gov/electricity/data/browser/#/topic/7?agg=2,0,1&geo=g&freq=M%2013:41:41%20GMT-0500%20(EST)){.uri}]. But this week we will work with the all.sectors column instead of residential price.

```{r}

#Importing time series data from text file#
electricity_price <- read.csv(
  file="./Data/Average_retail_price_of_electricity_United_States_monthly.csv",
  header=TRUE,
  skip=4)

#Inspect data
#head(electricity_price)
nvar <- ncol(electricity_price) - 1
nobs <- nrow(electricity_price)

#Preparing the data - create date object and rename columns
electricity_price_processed <-
  electricity_price %>%
  mutate( Month = my(Month) ) %>% 
  rename( All.sectors = all.sectors.cents.per.kilowatthour ) %>% 
  rename( Residential = residential.cents.per.kilowatthour ) %>% 
  rename( Commercial = commercial.cents.per.kilowatthour ) %>% 
  rename( Industrial = industrial.cents.per.kilowatthour ) %>% 
  arrange( Month )

#head(electricity_price_processed)
#summary(electricity_price_processed)

#No NAs so we don't need to worry about missing values

```

## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*.

```{r}
first_m <- month(electricity_price_processed$Month[1])
first_y <- year(electricity_price_processed$Month[1])
ts_electricity_price <- ts(
  electricity_price_processed[,2:(nvar+1)],
  start=c(first_y,first_m),
  frequency=12) 

#note that we are only transforming columns with electricity price, not the date columns  
#head(ts_electricity_price,15)
#tail(ts_electricity_price,15)

```

## Initial Plots

```{r}
TS_Plot <- 
  ggplot(electricity_price_processed, aes(x=Month, y=All.sectors)) +
      geom_line()
plot(TS_Plot)

#ACF and PACF plots
par(mfrow=c(1,2))
ACF_Plot <- Acf(electricity_price_processed$All.sectors, lag = 40, plot = TRUE,main="")
PACF_Plot <- Pacf(electricity_price_processed$All.sectors, lag = 40, plot = TRUE,main="")
par(mfrow=c(1,1))
```

## Decomposing the time series

The plots from the previous section show the data has a seasonal component. Since we are working with non-seasonal ARIMA, we need to decompose the series and eliminate the seasonality.

```{r}
#Using R decompose function
decompose_allsectors_price <- decompose(ts_electricity_price[,"All.sectors"],"additive")
plot(decompose_allsectors_price)

#The ACF plot show a slow decay which is a sign of non-stationarity.
#Creating non-seasonal residential price time series because some models can't handle seasonality
deseasonal_allsectors_price <- seasadj(decompose_allsectors_price)  
```

## Fitting Models to the original (seasonal) series

This week the goal is to expand the list of model we had for M8 and M9 and check accuracy of all the models. Let's work with out-of-sample forecasting, i.e., forecast observations that are not in the sample used to fit the models. There is a simple way to do that by setting argument `holdout=TRUE`. But some functions to do not offer the holdout argument. Therefore let's create a new series that will be a subset of our original one except that the last observation will be December 2019. And then we will store the remaining observations in 2020 to check the accuracy of our model forecasting 11 steps ahead of time.

```{r}
# Create a variable with number of steps ahead we will forecast
# This is helpful whenever you need to change forecasting horizon you only need to change it here
n_for <- 11 #play with this, change to 12, 23, 2 and check of the best model is still the same

ts_price <- ts(ts_electricity_price[1:(nobs-n_for),"All.sectors"],
               start=c(first_y,first_m),
               frequency=12)
last_obs <- ts_electricity_price[(nobs-n_for+1):nobs,"All.sectors"] 
#no need to make obs_2020 a ts object

#If you not sure what we did here comment the following lines and check the values/dates
#tail(ts_price,12)
#last_obs
```



We will use these new series to fit/forecast all the models. Do not forget to set holdout=FALSE to make sure we are forecasting the same time period. Let's start by looking at residual plots and AIC to check how the models represent the historical prices.

### Model 1: Arithmetic mean
```{r}
# The meanf() has no holdout option
MEAN_seas <- meanf(y = ts_price, h = n_for)  
checkresiduals(MEAN_seas)
plot(MEAN_seas)
```
Note a clear trend on residuals series, showing that the mean is not a good to model the trend component. And aside from trend the seasonal component is also not being modeled.

### Model 2: Seasonal naive
```{r}
SNAIVE_seas <- snaive(ts_price, h=n_for, holdout=FALSE)
checkresiduals(SNAIVE_seas)
plot(SNAIVE_seas)
```
The residuals for the seasonal naive don't seem to have a strong trend. Because it repeats the observations that happen in a previous seasonal lag (in this case one year ago), the seasonal naive is able to model the trend and seasonal component. But the residuals series show a strong autoregressive component which is also not desired.

### Model 3: SARIMA
```{r}
SARIMA_autofit <- auto.arima(ts_price)
checkresiduals(SARIMA_autofit)

#Generating forecasts
#remember auto.arima does not call the forecast() internally so we need one more step
SARIMA_for <- forecast(SARIMA_autofit,h=n_for)
plot(SARIMA_for)
```
Notice the residual series seems to be random and ACF shows no significant self correlation.

## Fitting State Space Models to the original (seasonal) series

### Model 4: SS Exponential smoothing

Let's start by trying an exponential smoothing under the state space model. In teh videos we went over the state equation for level only. But the ES can be used to represent trend and seasonal as well, these are know as ETS (Exponential Trend and Seasonal) models. We will talk in more details on M11.

```{r}
SSES_seas <- es(ts_price,model="ZZZ",h=n_for,holdout=FALSE,silent="none")
plot(SSES_seas)
checkresiduals(SSES_seas)
```
Note that the best model here was a ETS(AAM) which means we are fitting an additive random error, an additive trend and a multiplicative seasonal component.

### Model 5: SS with StructTS()

Now let's try to fit one of the three SS models we learned: local level, local linear trend and the basic structure model (BSM). Since we have seasonality we need to go with the BSM. We will use the `StructTS()` function from package `forecast`.
The function takes a few arguments, here are the important ones:

**x**  time series object
**type** type of the model. If `type ='level'` R will fit a local level model. If `type ='trend'` R will fit a local linear trend model and if `type ='BSM'` R will fit a basic structural model. If argument is omitted, a BSM is used for a time series with frequency > 1, and a local trend model otherwise.
**fixed** optional numeric vector with values for the components variances in the following order $fixed=c(\sigma^2_\eta,\sigma^2_\xi,\sigma^2_\omega,\sigma^2_\epsilon)$. The first variance is with respect to level, second is trend, third is seasonal and fourth is observation. If supplied, only NA entries in fixed will be varied. This argument is useful for setting variances to zero.


```{r}
SS_seas <- StructTS(ts_price,
                    type="BSM",fixed=c(0,0.001,0.3,NA))   #this function has convergence issues
checkresiduals(SS_seas)

#Generating forecasts
# StructTS() does not call the forecast() internally so we need one more step
SS_for <- forecast(SS_seas,h=n_for)
plot(SS_for)
```
## Model Performance for forecasting 12 steps ahead

We are done with backward-looking assessment. Let's move on to the forward-looking assessment for all models.

### Function accuracy() from package `forecast`

The function accuracy() will return performance measures. It takes the main arguments:

**object** object of class forecast, or numerical values containing forecasts. **x** numerical vector containing observed values (optional).

If **x** is not provided the function will return performance measures for trainign set, i.e., based on historical data it will compare observed and fitted values.

The measures calculated are:

**ME:** Mean Error **RMSE:** Root Mean Squared Error **MAE:** Mean Absolute Error **MPE:** Mean Percentage Error **MAPE:** Mean Absolute Percentage Error **MASE:** Mean Absolute Scaled Error **ACF1:** Autocorrelation of errors at lag 1

### Checking accuracy of the five models

We will change this part a bit compared to M9 because I want the accuracy just for the forecasting window of 12 months. Just to simplify the code I will create a vector with the last 12 observations. At this point make sure that for all instances that when you forecast you specify `holdout=TRUE` because we want to access the accuracy of the forecast out-of-sample.

```{r}
#Model 1: Arithmetic mean
MEAN_scores <- accuracy(MEAN_seas$mean,last_obs)  #store the performance metrics

#Model 2: Seasonal naive 
SNAIVE_scores <- accuracy(SNAIVE_seas$mean,last_obs)

# Model 3:  SARIMA 
SARIMA_scores <- accuracy(SARIMA_for$mean,last_obs)

# Model 4:  SSES
SSES_scores <- accuracy(SSES_seas$forecast,last_obs)

# Model 5:  BSM 
SS_scores <- accuracy(SS_for$mean,last_obs)

```

### Compare performance metrics

Now we will create a data frame that combines performance metrics for all the three models. You can choose one metric to help you choose among models. For example let's say we want the model with lowest RMSE.

````{r}
#create data frame
seas_scores <- as.data.frame(rbind(MEAN_scores, SNAIVE_scores, SARIMA_scores,SSES_scores,SS_scores))
row.names(seas_scores) <- c("MEAN", "SNAIVE","SARIMA","SSES","BSM")

#choose model with lowest RMSE
best_model_index <- which.min(seas_scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(seas_scores[best_model_index,]))                       
                            
```

SARIMA was the best fit for the seasonal data. If you want generate a table to compare model accuracy and help visualize the results here is a suggestion on how to include a table on your Rmd report. You can use the `kable_styling(latex_options="striped")` to highlight the model that leads to minimum RMSE.

```{r}


kbl(seas_scores, 
      caption = "Forecast Accuracy for Seasonal Data",
      digits = array(5,ncol(seas_scores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(seas_scores[,"RMSE"]))
```
## Plotting everything together

Here we will use autoplot() and autolayer() from package `ggplot2` to draw a particular plot for time series. The function autolayer() takes a few main arguments.

**x** Forecast object produced by forecast() function. If forecasts were generated with another function you may need to point to the object either mean or forecast to get the values .
**include** number of values from time series to include in plot. Default is all values.
**PI** Logical flag indicating whether to plot prediction intervals.
**series** Matches an unidentified forecast layer with a coloured object on the plot.

```{r}

autoplot(ts_electricity_price[,"All.sectors"]) +
  autolayer(MEAN_seas, PI=FALSE, series="Mean") +
  autolayer(SNAIVE_seas, PI=FALSE, series="Naïve") +
  autolayer(SARIMA_for,PI=FALSE, series="SARIMA") +
  autolayer(SSES_seas$forecast, series="SSES") +
  autolayer(SS_for,PI=FALSE,series="BSM") + 
  xlab("Month") + ylab("Electricity Retail Price ($/kWh)") +
  guides(colour=guide_legend(title="Forecast"))
```

