---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 6 - Due date 03/26/21"
author: "Marie McNamara" 
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp21.Rmd"). Submit this pdf using Sakai.

## Set up

```{r}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(sarima)
```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r}
natural_gas_new <- read.csv(file="../Data/Net_generation_United_States_all_sectors_monthly.csv" ,header=TRUE,skip = 4) 

first <- 
  natural_gas_new %>%
  select("Month","natural.gas.thousand.megawatthours") %>% 
  mutate( Month = my(Month)) %>% 
  rename( natural_gas = natural.gas.thousand.megawatthours)  %>% 
  arrange (Month)


nvar= ncol(first) - 1
ts_nat_price <- ts(first[,2:(nvar+1)],
                           start=c(year(first$Month[1]),month(first$Month[1])),
                           frequency=12) 

acf1 = Acf(ts_nat_price, lag.max = 40, plot = TRUE) 
pacf1 = Pacf(ts_nat_price, lag.max = 40, plot = TRUE) 

```

### Q2

Using the $decompose()$ or $stl()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the depersonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.


Based on the PACF and ACF plots there is a clear seasonal component. There also was an observed decreasing mean trend. The desasoned plot confirms that there is a decreasing seasonal component, as well as sesonality. 

This can be seen by the decreasing mean over time, and have a high correlation component between lags.
```{r}

natural_gas_deseas <- stl(ts_first[,2], s.window = "period") 
plot(natural_gas_deseas)

decompose_natural_price <- decompose(ts_first[,"natural_gas"],"additive")
plot(decompose_natural_price)
```

## Modeling the seasonally adjusted or deseasonalized series

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

The seasonal Mann kendall tests for P value was < 2.22e-16 which is less than 0.05. I failed to reject the null, that natural gas data is stationary, thus the data is not stationary and has a stocastic trend  

For the ADF test p=0.01 greater than 0.05. I rejected the null 
```{r}

i=2 
SMKtest <- SeasonalMannKendall(ts_first[,i])
print(summary(SMKtest))

i=2
print(adf.test(ts_first[,i],alternative = "stationary"))
```

### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to can read the plots and interpret the test results.

-Ask about removing seasonal component? 


After analyzing the ACF and PACF plots I notice that there is slow decay in the ACF Plot. There is also a PACF cut off at lag one which leds me to conclude that this is a AR process order 1. 

The series needs differencing as the Mankendal test above showed a stocastic trend. after running the n_diff funciton think there is only one differncing needed. the ACF plot determines q 

p=1
d=1
q=0

```{r}
deseasonal_natural<- seasadj(decompose_natural_price)
n_diff <- ndiffs(deseasonal_natural)
cat("Number of differencing needed:", n_diff)


```

### Q5

Use $Arima()$ from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. Should you allow for constants in the model, i.e., $include.mean = TRUE$ or $include.drift=TRUE$. **Print the coefficients** in your report. Hint: use the $cat()$ function to print.

```{r}

SARIMA_manual <- Arima(ts_first[,2],order=c(1,1,0), include.mean = TRUE,include.drift=TRUE)
print(SARIMA_manual)
cat("Coefficients", var.coef) 

```

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF and PACF on the same window. You may use the $checkresiduals()$ function to automatically generate the three plots. Do the residual series look like a white noise series? Why?

## Modeling the original series (with seasonality)

### Q7

Repeat Q4-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.

```{r}

```


### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.

```{r}

```


## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the correct orders. The intention of the assignment is to walk you to the process and help you figure out what you did wrong (if you did anything wrong!).

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}

```


### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}

```

