---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 5 - Due date 03/12/21"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.\
Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A05_Sp21.Rmd"). Submit this pdf using Sakai.

## Questions

This assignment has general questions about ARIMA Models.

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(forecast)
library(tseries)
```

## Q1

Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models:

\begin{enumerate}[label=(\alph*)]

\item AR(2)

> Answer: p=2. The ACF plot of an AR(2) model will have a slow decay and the PACF plot will have a clear cut off at lag 2.

\item MA(1)

> Answer: q=1. The PACF plot of a MA(1) model will have a slow decay and the ACF plot will have a clear cut off at lag 1. 

\end{enumerate}

## Q2

Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models.

```{r}
#ARMA(1,0)
ARMAmodel_1<- arima.sim(model=list(ar=0.6), n=100) #the AR coefficient is 0.6
#ARMA(0,1)
ARMAmodel_2<- arima.sim(model=list(ma=0.9), n=100) #the MA coefficient is 0.9
#ARMA(1,1)
ARMAmodel_3<- arima.sim(model=list(ar=0.6,ma=0.9), n=100) #the AR coefficient is 0.6 and the MA coefficient is 0.9

#plot the time series, optional 
par(mfrow=c(1,3))
plot(ARMAmodel_1, main="Time Series of ARMA(1,0)")
plot(ARMAmodel_2, main="Time Series of ARMA(0,1)")
plot(ARMAmodel_3, main="Time Series of ARMA(1,1)")
```


\begin{enumerate}[label=(\alph*)]

\item Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r}
#plot the ACF
par(mfrow=c(1,3))
acf_ARMAmodel_1 <- Acf(ARMAmodel_1, type="correlation", plot=T)
acf_ARMAmodel_2 <- Acf(ARMAmodel_2, type="correlation", plot=T)
acf_ARMAmodel_3 <- Acf(ARMAmodel_3, type="correlation", plot=T)


```


\item Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r}
#plot the PACF
par(mfrow=c(1,3))
pacf_ARMAmodel_1 <- Acf(ARMAmodel_1, type="partial", plot=T)
pacf_ARMAmodel_2 <- Acf(ARMAmodel_2, type="partial", plot=T)
pacf_ARMAmodel_3 <- Acf(ARMAmodel_3, type="partial", plot=T)
```

\item Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: It's relatively easy to tell model 1 is a autoregressive model with p=1 because of a slow decay in the ACF plot and a clear cut off at lag 1 in the PACF plot (p=1). For model 2, you can tell it's a moving average model because the PACF plot have a slow decay and the ACF plot has a cut off at lag 1 (q=1). For model 3, it's hard to tell because we are superimposing AR and MA properties. 

\item Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

```{r}
print(acf_ARMAmodel_1) 
print(pacf_ARMAmodel_1) 

print(acf_ARMAmodel_2) 
print(pacf_ARMAmodel_2) 

print(acf_ARMAmodel_3)
print(pacf_ARMAmodel_3) 
```

Answer: For model 1 and 3, since the order of AR component is 1, the value of PACF at lag 1 is close to the value we specify. For model 2 with only MA process, you can't read the coefficient value on ACF or PACF. 


\item Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r}
#ARMA(1,0)
ARMAmodel_1_e<- arima.sim(model=list(ar=0.6), n=1000) #the AR coefficient is 0.6
#ARMA(0,1)
ARMAmodel_2_e<- arima.sim(model=list(ma=0.9), n=1000) #the MA coefficient is 0.9
#ARMA(1,1)
ARMAmodel_3_e<- arima.sim(model=list(ar=0.6,ma=0.9), n=1000) #the AR coefficient is 0.6 and the MA coefficient is 0.9

par(mfrow=c(2,3))
acf_ARMAmodel_1_e <- Acf(ARMAmodel_1_e, type="correlation", plot=T)
acf_ARMAmodel_2_e <- Acf(ARMAmodel_2_e, type="correlation", plot=T)
acf_ARMAmodel_3_e <- Acf(ARMAmodel_3_e, type="correlation", plot=T)
pacf_ARMAmodel_1_e <- Acf(ARMAmodel_1_e, type="partial", plot=T)
pacf_ARMAmodel_2_e <- Acf(ARMAmodel_2_e, type="partial", plot=T)
pacf_ARMAmodel_3_e <- Acf(ARMAmodel_3_e, type="partial", plot=T)
```

Answer: The diagnostic capabilities of these plots are higher with increasing number of observations. It's easier to tell model 1 is an autoregressive model with p=1, and model 2 is a moving average model with q=1. And model 3, the ARMA model, is a superposition of the AR and MA model. 

\end{enumerate}

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

\begin{enumerate}[label=(\alph*)]

\item Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

> Answer: From the model equation, yt depends in two previous observations, yt-1 and yt-12, since 12 is our seasonal lag the order of the AR component will be p=1 and SAR component will be P=1. The value of yt also depends on the previous residual at-1 meaning we have a MA component of order q=1. There is no a_t-12 so SMA component order will be Q=0.

> With respect to the number of differencing it’s hard to tell from the equation but there are two things you could argue. The first is there is no constant term, therefore we are either working with a zero-mean process or this series had already been differenced,leading to the conclusion that either d=1 or D=1. 

Thus, this equation represents ARIMA(1,d,1)(1,D,0)_12.  

\item Also from the equation what are the values of the parameters, i.e., model coefficients. 

phi_1=0.7, phi_12=-0.25, theta=-0.1

\end{enumerate}


## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)(1, 0)_{12}$ model with $\phi =0.8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
#install.packages("sarima")
library(sarima)
model_4<- sim_sarima(model=list(ma=0.5,sar=0.8, nseasons=12), n=1000) 
#moving average coefficient for nonseasonal component is 0.5; the AR coefficient for seasonal component is 0.8. 
#could also use sarima.sim() from package "astsa"
#ACF and PACF 
par(mfrow=c(1,2))
Acf(model_4, type="correlation", plot=T,lag.max=60)
Acf(model_4, type="partial", plot=T,lag.max=60)

#coefficients
coeffi_model_4=Arima(model_4, order=c(0,0,1), seasonal=list(order=c(1,0,0), period=12))
print(coeffi_model_4)
```

> Answer: For the order of the non-seasonal component, we focus on the early lags (ex.before lag=12). We can see that PACF shows a slow decay and ACF has a clear cut off at lag 1, which indicates that the non-seasonal component has moving average process, q=1.

> As for the seasonal component, we are only intrested at seasonal lags 12, 24, 36 and so forth. We can see that the ACF has multiple spikes at the seasonal lags and PACF has only one spike at lag 12. It shows that we should include a SAR process, P=1.

> The plot result is consistent with the parameters specified in the simulation. Note that the PACF at lag 12 is 0.7, which is close to the AR coefficient phi_12=0.8. The ACF at lag 1 is around 0.4, which is close to the MA coefficient theta_1=0.5. The coefficients of the simulated series are very close to the theorectical coefficients.
